Type-Safe Modular Parsing

### 2017-03-21 paper/response.txt

We thank all reviewers for their thoughtful comments. Below we would like to talk about our contributions, and
some questions raised by reviewers.

As we discussed in the paper, almost all previous attempts on extensible/modular parsing stop at syntactical
level of modularity for various reasons. We propose a pattern that nicely achieves semantic modularity and
simplicity for parsing by using standard OO techniques. It’s a notable improvement, since compiling and
generating the parser again during extension every time is obviously a worse practice.

Our framework is indeed based on existing techniques, because a concrete algorithm is required for parsing.
In particular, we use Packrat (PEG) parsing and Object Algebras in our Scala implementation. However, we show
in our case study that they integrate well as expected; the parsing code is elegant and simplified to a great
extent, as boilerplate would explode along with a series of compositions. As discussed in Section 5.4 of the
paper, our approach can also be generalized thus not limited to any specific techniques. What we want to propose
is really a simple, powerful and practical framework, or a flexible platform, instead of a library, to achieve
easy and type-safe composition.

We are sorry that we missed the material of paper [1], as modular parsing and the pattern is not explained
in detail in the paper. Their code online seems to be different with ours as well. Based on parser
combinators, our pattern is more concise and easy to use.

Language components are an important contribution of our work, where we encapsulate a language feature with
three parts: abstract syntax, parser, and operations; everything is modular. With semantically modular parsing,
such components are fully type-safe. Just as SugarJ work and so on, we want languages to be small components,
that can be composed in a flexible way into modular languages again. They are reusable for further composition.
We don't need to rely on a base language to perform extensions as many extensible compilers do.

Rev B:
> However, object algebras solve a problem that is orthogonal to parsing...

We think Object Algebras are not a total orthogonal direction from modular parsing.
Since for parsing, the extension could happen in two cases:
-    When syntax changed, the parser should be extended easily to cooperate with the new syntax.
-    When ASTs changed (a new data variant/new operation added), the old parser should produce
     compatible ASTs without modification.
Regarding the second type of extension, Object Algebras or similar techniques are needed for
modular parsing. Our intuition of using Object Algebras, is not only it encodes multiple syntax
nicely, but also it adds a different dimension to obtain full extensibility of parsing,
supporting separate compilation.

Rev C:
> I would have liked to see it used for more realistic programming language grammars.

We agree that it would be better to show a more realistic language for demonstration. Thanks for your suggestion.

Rev C:
> - Unclear to me if indirect left recursion is supported, which would be needed
> for practical grammars.

Packrat parsing theoretically supports both direct and indirect left recursion [2],
although the implementation in Scala’s standard parser combinator library may be buggy
for some indirect left recursion cases.

Rev C:
> - Does not provide a sufficiently complete solution to language componentization,
> since object algebras are too restricted (only support bottom up computations).

Indeed Object Algebras only support bottom-up traversals, but we have known that some
other researchers have worked out some framework that transforms Object Algebra ASTs
to different ones, which also supports top-down traversals.

Rev E:
> Why is this "type-safety" of the parser so important, after all?  Yes, one may have
> bugs or missing cases in the parser itself, but I would not expect that to be a
> problem, because one should have tests that cover all the grammar anyway, so it is
> little likely that the type-safety of the parser itself would bring much practical benefits.

We think type-safe composition and separate compilation really matter, since it is not a good
idea to modify existing code; we are using OO and want to achieve better code reuse. As mentioned
before, we hope project code to be modular and easily composable, instead of being compiled and
generating code for every round of composition. It is better for debugging.

[1]: Rendel, Tillmann, Jonathan Immanuel Brachthäuser, and Klaus Ostermann. From object algebras to attribute grammars. OOPSLA 2014.
[2]: Alessandro Warth, James R Douglass, and Todd D Millstein. Packrat parsers can support left recursion. PEPM 2008.

### 2017-03-23 paper/response.txt

We thank all reviewers for their thoughtful comments. Below we would
like to talk about our contributions, and some questions raised by
reviewers.

The essential criticism seems to be well summarized by reviewer D:

"Object algebras can be used to modularize tree traversals. Parser
combinators combined with a Packrat algorithm and open recursion allow
modular parser definitions. To people that know both techniques,
object algebras and Packrat parsing, it also appears obvious to
combine the two."

We should also add, on top of the 2 techniques above, there's a third
thing: *good knowledge of some of Scala's more advanced features.*

Is it true that the paper is essentially combining 2 techniques + good
use of Scala to achieve modular parsing, and that people that know
about all of this could possibly recreate this work?

Yes, it is true.

Is that a bad thing?

We are not sure about this. There are at least two arguments
for why the fact that the technique actually ends up being quite
simple is actually a good thing:

1) *It shows that semantically modular parsing can be
seriously considered for practical use*.

Semantic modularization techniques are still far behind syntactic
modularization techniques in terms of proving their
worth/applicability in practice. Without question syntactic modularization
approaches such as those currently used by Language Workbenches,
are currently much more well-accepted in practice. This is in part
because semantic modularization techniques are underdeveloped, and
in part because there is an "established reputation" that semantic
modularization techniques are just simply too hard to use in practice
(and don't scale well).

We believe the simplicity in our work/techniques makes a good
argument for semantic modularization techniques.

2) *We would dare to say that the set of people that knows
Packrat parsing, Object Algebras and Scala well-enough to
know that type-safe modular parsing can be easily done
(before seeing our paper!) is less than 5*

Honestly we didn't know this when we started our work (so we were
not in that set). Although we knew Object Algebras and Scala
well, we were not well-versed with Packrat parsing. We spent quite
a bit of time playing with other parsing combinator libraries, and
getting into trouble before we tried Packrat parsing and solved the
issues we encountered.

Finally, language components and our case study are important
contributions of our work. Language components
encapsulate a language feature with three parts:
abstract syntax, parser, and operations; everything is modular. With
semantically modular parsing, such components are fully
type-safe. Just as SugarJ work and so on, we want languages to be
small components, that can be composed in a flexible way into modular
languages again. They are reusable for further composition.  We don't
need to rely on a base language to perform extensions as many
extensible compilers do. Our case study is also a good benchmark
for other researchers and we believe a good contribution of this work.

@@

Rev B:
> However, object algebras solve a problem that is orthogonal to parsing...

If the goal is to modularize languages (both parsing and semantics),
and have libraries of language components, we don't see how the issues
are orthogonal. The techniques must be combined.

@@

Rev C:
> I would have liked to see it used for more realistic programming language grammars.

We agree that it would be better to show a more realistic language for
demonstration. Thanks for your suggestion.

Rev C:
> - Unclear to me if indirect left recursion is supported, which would be needed
> for practical grammars.

Packrat parsing theoretically supports both direct and indirect left recursion [2],
although the implementation in Scala's standard parser combinator library may be buggy
for some indirect left recursion cases.

@@

Rev C:
> - Does not provide a sufficiently complete solution to language componentization,
> since object algebras are too restricted (only support bottom up computations).

Indeed Object Algebras only support bottom-up traversals, but there's
ongoing work on variants of Object Algebras that remove this
limitation. We think our approach will work just fine with such
variants.

Rev D:

We are sorry that we missed the material of paper [1], as modular
parsing and the pattern is not explained in detail in the paper. Their
code online seems to be different with ours as well. Based on parser
combinators, our pattern seems to be more concise and simpler to use.

### 2017-04-24 ECOOP_17/paper.tex
Move ‘paper’ to ‘ECOOP_17’

  This paper investigates solutions for the problem of \emph{modular
    parsing}. We focus on \emph{semantic} modularity and not just
  \emph{syntactic} modularity. That is, the solutions should not only
  allow complete parsers to be built out of modular parsing
  components, but also enable the parsing components to be
  \emph{modularly type-checked} and \emph{separately compiled}. We
  present a technique based on parsing combinators that enables modular
  parsing. Interestingly, the modularity requirements for modular
  parsing rule out several existing parsing combinator approaches,
  which rely on some non-modular techniques. We show that Packrat
  parsing techniques, provide solutions for such modularity problems,
  and enable reasonable performance in a modular setting.
  Extensibility is achieved using \emph{multiple inheritance} and
  \emph{Object Algebras}.  To evaluate
  the approach we conduct a case study based on the ``Types and
  Programming Languages'' interpreters. The case study shows the
  effectiveness at reusing parsing code from existing interpreters,
  and the total parsing code is 69\% shorter than an existing code
  base using non-modular parsing code.

>>>>

  This paper investigates solutions for the problem of \emph{modular
    parsing}. We focus on \emph{semantic} modularity and not just
  \emph{syntactic} modularity. That is, the solutions should not only
  allow complete parsers to be built out of modular parsing
  components, but also enable the parsing components to be
  \emph{modularly type-checked} and \emph{separately compiled}. We
  present a technique based on parser combinators that enables modular
  parsing. Interestingly, the modularity requirements for modular
  parsing rule out several existing parser combinator approaches,
  which rely on some non-modular techniques. We show that Packrat
  parsing techniques, provide solutions for such modularity problems,
  and enable reasonable performance in a modular setting.
  Extensibility is achieved using \emph{multiple inheritance} and
  \emph{Object Algebras}.  To evaluate
  the approach we conduct a case study based on the ``Types and
  Programming Languages'' interpreters. The case study shows the
  effectiveness at reusing parsing code from existing interpreters,
  and the total parsing code is 69\% shorter than an existing code
  base using a non-modular parsing approach.

### 2017-04-24 ECOOP_17/sections/CaseStudy.tex
Move ‘paper’ to ‘ECOOP_17’

\begin{table}
    \centering
    \begin{tabular}{|l|r|r|r|r|r|r|r|}
      \hline
        \multirow{2}{*}{\bfseries Calculus Name} & \ilyaimpl{} & \multicolumn{2}{ c| }{\ourimpl{}} & \multicolumn{2}{ c| }{\ilyalongest{}} & \multicolumn{2}{ c| }{\ourclass{}} \\ \cline{2-8}
        \multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{\bfseries Time} & \bfseries Time & \bfseries (+/-)\% & \bfseries Time & \bfseries (+/-)\% & \bfseries Time & \bfseries (+/-)\% \\
      \hline
        Arith & 741 & 913 & +23.2 & 793 & +7.0 & 932 & +25.8 \\
        Untyped & 770 & 1018 & +32.2 & 821 & +6.6 & 1007 & +30.8 \\
      \hline
        \multicolumn{1}{|c|}{\dots} & \multicolumn{7}{c|}{\dots} \\
      \hline
        FullRecon & 1094 & 1645 & +50.4 & 1161 & +6.1 & 1652 & +51.0 \\
        FullPoly & 1398 & 2086 & +49.2 & 1511 & +8.1 & 2019 & +44.4 \\
        FullOmega & 1451 & 2352 & +62.1 & 1582 & +9.0 & 2308 & +59.1 \\
      \hline
        Total & 21746 & 31024 & +42.7 & 23260 & +7.0 & 30795 & +41.6 \\
      \hline
        \multicolumn{8}{c}{}
    \end{tabular}
    \caption{Execution time of four implementations.}
    \label{tab:ext-comparison}
\end{table}

Table \ref{tab:ext-comparison} shows part of the comparison results of these four versions,
full table is available online. It suggests that the difference of running time between
using object algebra and class is little, roughly 1\%.
The usage of longest match combinator slows the performance by 7\%. The main reason of slower
execution time may be the overall structure of the modular parsing approach, because we indeed have
more intermediate function calls and method overriding. However, it is worth mentioning that
because of the memoization technique of Packrat parsers, we are only constant times
slower, the algorithmic complexity is still the same.

>>>>

\begin{table}
    \centering
    \begin{tabular}{|l|r|r|r|r|r|r|r|}
      \hline
        \multirow{2}{*}{\bfseries Calculus Name} & \ilyaimpl{} & \multicolumn{2}{ c| }{\ourimpl{}} & \multicolumn{2}{ c| }{\ilyalongest{}} & \multicolumn{2}{ c| }{\ourclass{}} \\ \cline{2-8}
        \multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{\bfseries Time} & \bfseries Time & \bfseries (+/-)\% & \bfseries Time & \bfseries (+/-)\% & \bfseries Time & \bfseries (+/-)\% \\
      \hline
        Arith & 741 & 913 & +23.2 & 793 & +7.0 & 932 & +25.8 \\
        Untyped & 770 & 1018 & +32.2 & 821 & +6.6 & 1007 & +30.8 \\
        FullUntyped & 1297 & 1854 & +42.9 & 1343 & +3.5 & 1767 & +36.2 \\
        TyArith & 746 & 888 & +19.0 & 772 & +3.5 & 918 & +23.1 \\
        SimpleBool & 1376 & 1782 & +29.5 & 1494 & +8.6 & 1824 & +32.6 \\
        FullSimple & 1441 & 2270 & +57.5 & 1574 & +9.2 & 2226 & +54.5 \\
        Bot & 1080 & 1287 & +19.2 & 1078 & -0.2 & 1306 & +20.9 \\
      %\hline
        %\multicolumn{1}{|c|}{\dots} & \multicolumn{7}{c|}{\dots} \\
      %\hline
        FullRef & 1438 & 2291 & +59.3 & 1544 & +7.4 & 2142 & +49.0 \\
        FullError & 1410 & 1946 & +38.0 & 1524 & +8.1 & 1981 & +40.5 \\
        RcdSubBot & 1247 & 1524 & +22.2 & 1285 & +3.0 & 1612 & +29.3 \\
        FullSub & 1320 & 1979 & +49.9 & 1393 & +5.5 & 1899 & +43.9 \\
        FullEquiRec & 1407 & 2200 & +56.4 & 1561 & +10.9 & 2156 & +53.2 \\
        FullIsoRec & 1492 & 2253 & +51.0 & 1648 & +10.5 & 2236 & +49.9 \\
        EquiRec & 994 & 1254 & +26.2 & 1048 & +5.4 & 1304 & +31.2 \\
        Recon & 1044 & 1482 & +42.0 & 1128 & +8.0 & 1506 & +44.3 \\
        FullRecon & 1094 & 1645 & +50.4 & 1161 & +6.1 & 1652 & +51.0 \\
        FullPoly & 1398 & 2086 & +49.2 & 1511 & +8.1 & 2019 & +44.4 \\
        FullOmega & 1451 & 2352 & +62.1 & 1582 & +9.0 & 2308 & +59.1 \\
      \hline
        Total & 21746 & 31024 & +42.7 & 23260 & +7.0 & 30795 & +41.6 \\
      \hline
        \multicolumn{8}{c}{}
    \end{tabular}
    \caption{Execution time of four implementations.}
    \label{tab:ext-comparison}
\end{table}

Table \ref{tab:ext-comparison} shows comparison results of these four versions. It suggests that the difference of running time between
using object algebra and class is little, roughly 1\%.
The usage of longest match combinator slows the performance by 7\%. The main reason of slower
execution time may be the overall structure of the modular parsing approach, because we indeed have
more intermediate function calls and method overriding. However, it is worth mentioning that
because of the memoization technique of Packrat parsers, we are only constant times
slower, the algorithmic complexity is still the same.

### 2017-04-24 ECOOP_17/sections/Introduction.tex
Move ‘paper’ to ‘ECOOP_17’

\paragraph{Algorithmic Challenges} A first challenge was to do with
  the parsing algorithms themselves. Since many parsing formalisms
  were not designed with extensibility in mind, some of the employed
  techniques were non-modular, or performed poorly in a modular
  setting.  The most widely use tools for parsing are parser
  generators, but most algorithms employed by parser generators
  require full information about the grammar to generate parsing
  code. Moreover, actions associated to grammar productions are
  typically only type-checked after the parser has been
  generated. Both problems go against our goals of semantic
  modularity.

@@

  To address such algorithmic challanges, we propose a methodology
  for implementing modular parsers built on top of an
  existing Packrat~\cite{Ford2002} parsing library for Scala~\cite{odersky2004overview}. The Packrat parsing library
  directly supports left-recursion, and a \emph{longest-match
    composition} operator. The direct support for left-recursion
  avoids the need for left-recursion elimination, which is very
  cumbersome is a modular setting. The longest-match composition
  operator avoids the use of priorities, and it provides reasonably
  efficient optimizations for backtracking.

@@

\item {{\bf TAPL case study:}} We conduct a case study with 18 interpreters
  from the TAPL book. The case study show the effectiveness of modular
  parsing in terms of reuse. The TAPL case study is available online
  at:
  \url{http://}\footnote{{\bf Note to Reviewers:} Please find the url in the
  suplementary material for the submission.}

>>>>

\paragraph{Algorithmic Challenges} A first challenge was to do with
  the parsing algorithms themselves. Since many parsing formalisms
  were not designed with extensibility in mind, some of the employed
  techniques were non-modular, or performed poorly in a modular
  setting.  The most widely use tools for parsing are parser
  generators, but most algorithms employed by parser generators
  require full information about the grammar to generate parsing
  code. Moreover, actions associated with grammar productions are
  typically only type-checked after the parser has been
  generated. Both problems go against our goals of semantic
  modularity.

@@

  To address such algorithmic challenges, we propose a methodology
  for implementing modular parsers built on top of an
  existing Packrat~\cite{Ford2002} parsing library for Scala~\cite{odersky2004overview}. The Packrat parsing library
  directly supports left-recursion, and a \emph{longest-match
    composition} operator. The direct support for left-recursion
  avoids the need for left-recursion elimination, which is very
  cumbersome is a modular setting. The longest-match composition
  operator avoids the use of priorities, and it provides reasonably
  efficient optimizations for backtracking.

@@

\item {{\bf TAPL case study:}} We conduct a case study with 18 interpreters
  from the TAPL book. The case study show the effectiveness of modular
  parsing in terms of reuse. The TAPL case study is available online
  at:
  \url{http://}\footnote{{\bf Note to Reviewers:} Please find the URL in the
  supplementary material for the submission.}

### 2017-04-24 ECOOP_17/sections/PackratParsing.tex
Move ‘paper’ to ‘ECOOP_17’

This section discusses the algorithmic challenges introduced by
modular parsing and argues that Packrat parser combinators~\cite{Ford2002}
are suitable to address those challenges. The algorithmic challenges
are important because they rule out various common techniques
used by non-modular code using parsing combinators.
To avoid pitfalls related to those algorithmic challenges,
we propose the following methodology:

@@

After left-recursion elimination, the structure of grammar is changed,
as well as its corresponding parser. In a modular setting, it is
possible but unnecessarily complicated to analyse the grammar and
rewrite it when doing extensions. Anticipating that every non-terminal
has left-recursive rules is helpful for extensibility but overkill,
since it is inconvenient and introduces extra complexity for
representation of a grammar and implementation of a parser.

>>>>

This section discusses the algorithmic challenges introduced by
modular parsing and argues that Packrat parser combinators~\cite{Ford2002}
are suitable to address those challenges. The algorithmic challenges
are important because they rule out various common techniques
used by non-modular code using parser combinators.
To avoid pitfalls related to those algorithmic challenges,
we propose the following methodology:

@@

After left-recursion elimination, the structure of grammar is changed,
as well as its corresponding parser. In a modular setting, it is
possible but unnecessarily complicated to analyse the grammar and
rewrite it when doing extensions. Anticipating that every non-terminal
has left-recursive rules is helpful for extensibility but overkill,
since it is inconvenient and introduces extra complexity for
representation of grammars and implementation of parsers.
